Changes in Patterns of Geometric Sign Groupings as a Model of Social Dynamics Across Aurignacian Phases

### 2025-03-15 endofquarterpaper.qmd
stash

```{r}
#| label: tbl-aur-table
#| tbl-cap: "Summary of chronological, cultural and environmental characteristics of Aurignacian periods [@Banks_dErrico_Zilhao_2013;  @Shao_Limberg_Klein_Wegener_Schmidt_Weniger_Hense_Rostami_2021; @Chu_Richter_2020; @Fletcher_Sánchez]"
#| tbl-cap-location: bottom

@@

The first aim of our study was to determine how the grouping dynamics of geometric signs changed across the four phases of the Aurignacian, specifically in terms of diversity, size, connectivity, and boundaries. The results as displayed in @fig-seriation show that each phase can be divided into two main cultural groups, with diversity serving as a major dividing factor, and that group size varied from phase to phase, with a large increase in between the Transitional phase and the Proto-Aurignacian. @fig-neighbor-net shows that group inter-connectivity and inner-connectivity varies from phase to phase, with weak inner group connections and no between-group connections in the Transitional phase, to stronger inner connections and interconnections starting to form in the Proto-Aurignacian, to extremely strong connections (both between and within groups) in the Early Aurignacian, to weaker connections in the Evolved Aurignacian. The modularity tests, as shown in @tbl-strength-table, tell us that the groupings are strongest in the Transitional period and weakest in the Early Aurignacian, and the perMANOVA tests tell us that the Transitional period experiences the highest level of difference between groups, while the Early Aurignacian experiences the least. The Mantel tests, as shown in @tbl-mantel-table, tell us that the sign distribution in the Proto-Aurignacian experiences the lowest correlation with geographic distance, while the Early Aurignacian experiences the highest correlation.

@@

These results are supported by the existing data on the environmental and biological change established during these phases. @Shao_Wegener_Klein_Schmidt_Weniger_2024 tells us that from 45 - 43.25 k BP, the population density in Europe was extremely low, due to the unfavorable environmental conditions of 1Greenland Stadial 12 (44 k BP) and the slow westward expansion of early modern humans, before being followed by rapid expansion from 43.25 - 41 k BP, with the maximum extent of Aur-P1 settlement being reached around 41 k BP. This could provide an answer as to why the Transitional phase appears sparsely populated with little to no inter-connectivity, and the Proto-Aurignacian appears to be more densely populated with stronger cultural groups. Additionally @Banks_dErrico_Zilhao_2013 use eco-cultural niche modeling to show that the Heinrich Stadial 4 at the very beginning of the Early Aurignacian resulted in an expansion into new environmental niches, which they hypothesize was associated with expansion of social networks both within and between populations. This could be a reason as to why we see a large increase in both inter-connectivity and inner-connectivity strength in the Early Aurignacian, as well as the increased correlation with geographic distance. Lastly, @Shao_Limberg_Klein_Wegener_Schmidt_Weniger_Hense_Rostami_2021 developed a human-existence probability model for the Aurignacian, finding an expanded, more dispersed settlement area for Aur-P2, theorizing that the environmental changes of HS4 lead to humans better adapted to survive in a broader range of climate conditions. @Shao_Wegener_Klein_Schmidt_Weniger_2024 also find an increase in expansion marking the beginning of Aur-P2. This could explain why we see weaker connections in the Evolved Aurignacian, as expansion results in less interconnection.

>>>>

```{r}
#| label: tbl-aur-table
#| tbl-cap: "Table of data on Aurignacian periods [@Banks_dErrico_Zilhao_2013;  @Shao_Limberg_Klein_Wegener_Schmidt_Weniger_Hense_Rostami_2021; @Chu_Richter_2020; @Fletcher_Sánchez]"
#| tbl-cap-location: bottom

@@

The first aim of our study was to determine how the grouping dynamics of geometric signs changed across the four phases of the Aurignacian, specifically in terms of diversity, size, connectivity, and boundaries. The results as displayed in @fig-seriation show that each phase can be divided into two main cultural groups, with diversity serving as a major dividing factor, and that group size varied from phase to phase, with a large increase in between the Transitional phase and the Proto-Aurignacian.

@fig-neighbor-net shows that group inter-connectivity and intra-connectivity varies from phase to phase, with weak inner group connections and no between-group connections in the Transitional phase, to stronger intraconnections and interconnections starting to form in the Proto-Aurignacian, to extremely strong connections (both between and within groups) in the Early Aurignacian, to weaker connections in the Evolved Aurignacian. The modularity tests, as shown in @tbl-strength-table, show that the groupings are strongest in the Transitional period and weakest in the Early Aurignacian, and the perMANOVA tests tell us that the Transitional period experiences the highest level of difference between groups, while the Early Aurignacian experiences the least. The Mantel tests, as shown in @tbl-mantel-table, tell us that the sign distribution in the Proto-Aurignacian experiences the lowest correlation with geographic distance, while the Early Aurignacian experiences the highest correlation.

Map shows that geographic boundaries do not readily correspond to the group distributions and Mantel test results

@@

These results are supported by the existing data on the environmental and biological change established during these phases. @Shao_Wegener_Klein_Schmidt_Weniger_2024 tells us that from 45 - 43.25 k BP, the population density in Europe was extremely low, due to the unfavorable environmental conditions of Greenland Stadial 12 (44 k BP) and the slow westward expansion of early modern humans, before being followed by rapid expansion from 43.25 - 41 k BP, with the maximum extent of Proto/Early Aurignacian (Aur-P1) settlement being reached around 41 k BP. This rapid expansion could be what stimulated the emergence of sign sue as a marker of group membership.

Additionally @Banks_dErrico_Zilhao_2013 use eco-cultural niche modeling to show that the Heinrich Stadial 4 at the very beginning of the Early Aurignacian resulted in an expansion into new environmental niches, which they hypothesize was associated with expansion of social networks both within and between populations. This likely explains why we see a large increase in both inter-connectivity and inner-connectivity strength in the Early Aurignacian, as well as the increased correlation with geographic distance. Lastly, @Shao_Limberg_Klein_Wegener_Schmidt_Weniger_Hense_Rostami_2021 developed a human-existence probability model for the Aurignacian, finding an expanded, more dispersed settlement area for Evolved/Late Aurignacian (Aur-P2) theorizing that the environmental changes of HS4 lead to humans better adapted to survive in a broader range of climate conditions. @Shao_Wegener_Klein_Schmidt_Weniger_2024 also find an increase in expansion marking the beginning of the Evolved Aurignacian. Geographic expansion resulting in less group interconnection could explain why we see weaker connections in the Evolved Aurignacian.
